Testing for measurement invariance with respect to an ordinal variable

### 2013-04-15 ordinal/ordinal.Rnw
Added babel package to ordinal.Rnw, which combats the tilde problem in texi2pdf
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@195 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
small improvements, update of Rasch tree ref
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@196 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

@@

The test statistics proposed here are derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{KopZei10,MerZei13}.
In the following section, we provide an overview of the family and
describe the proposed statistics in detail.  Subsequently, we report on the
results of two simulation studies designed to compare the proposed
test statistics to existing tests of measurement
invariance.  Moreover, we illustrate the proposed statistics using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the tests' use in practice.

@@

As mentioned previously, when $V$ is categorical with a
relatively-small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
are related to those underlying Modification Indexes (e.g.,
\citeNP{Sor89}): the Modification Index is equivalent to a Lagrange
multiplier test, and the Lagrange multiplier test is contained in the
family described by \citeA{MerZei13}.  Here,
we focus on the tests' properties that are relevant for extending them
to the ordinal case.

@@

%% Paragraph on test statistics
Specific test statistics commonly obtained under this framework
include the double maximum statistic
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
which essentially tests whether any component of the cumulative score
process strays too far from zero and is easily visualized.  This test
discards information 
related to multiple parameters fluctuating simultaneously, resulting
in it having relatively-low power for assessing measurement
invariance when multiple factor analysis parameters change
simultaneously \cite{MerZei13}.

@@

The more interesting result of Figure~\ref{fig:sim2res} lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively-high power to
detect the measurement invariance violation, illustrating the result 
of \citeA{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.

@@

\subsection{Extension to Other Models}
We focused on testing for measurement invariance in factor
analysis models here, but the proposed test statistics are applicable to
other psychometric models that are estimated via ML (or similar estimation
techniques for independent observations that are governed by a central limit
theorem). The only
requirement for carrying out the tests is that the casewise scores
(Equation~\eqref{eq:score}) be available following model estimation.
As a result, applications to studying DIF in IRT are immediate (for a
presentation involving non-ordinal variables, see \citeNP{KopZei10}),
as are general psychometric applications to studying parameter
stability w.r.t.\ ordinal auxiliary variables.  These could 
include applications where ordinal variables are explicitly included
in the model, such as ordinal factor analysis.
We expect the
same general results to hold for these applications, whereby the
proposed test statistics are better than the LRT for detecting
monotonic instabilities.   The strucchange package \cite{Zei06} noted
previously can be used for these more-general applications.

\subsection{Summary}
As demonstrated via simulation, the proposed test statistics 
have relatively-high power for detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have relatively-low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous.  In all, the tests have
advantageous properties that should be useful in practice.

>>>>

\documentclass[man]{apa}
\usepackage{graphicx,epsfig,amsmath,alltt,setspace,bm}
%% This keeps the tildes out when you use Rnw2pdf:
\usepackage[english]{babel}
%% need no \usepackage{Sweave}
\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, results = hide}

@@

The test statistics proposed here are derived from a family of tests
that were recently applied to the study of measurement invariance in
psychometric models \cite{MerZei13,StrKop13}.
In the following section, we provide an overview of the family and
describe the proposed statistics in detail.  Subsequently, we report on the
results of two simulation studies designed to compare the proposed
test statistics to existing tests of measurement
invariance.  Moreover, we illustrate the proposed statistics using
psychometric data on scales purported to measure youth gratitude.  
Finally, we provide some detail on the tests' use in practice.

@@

As mentioned previously, when $V$ is categorical with a
relatively small number of categories, 
tests of measurement invariance typically proceed via multiple-group
models.  In this situation, we use likelihood ratio tests to compare a
model whose parameters differ across groups to a model whose
parameters are constrained to be equal across groups.  When $V$ is
continuous, however, multiple-group models usually cannot be used 
because there are no existing groups.  Instead, we can fit a model
whose parameters are restricted to be equal across all individuals and then
examine how individuals' scores ${\bm s}(\hat{{\bm \theta}}; x_i)$
fluctuate with their values of $V$.  If measurement invariance holds
with respect to $V$, then the scores should randomly fluctuate around
zero.  Conversely, if measurement invariance does not hold, then the
scores should systematically depart from zero.  These ideas are
related to those underlying the Lagrange multiplier test and are
discussed in detail by \citeA{MerZei13}.  Additionally, these ideas
are related to those underlying modification indexes (e.g.,
\citeNP{Sor89}): the modification index is equivalent to a Lagrange
multiplier test, and the Lagrange multiplier test is contained in the
family described by \citeA{MerZei13}.  Here,
we focus on the tests' properties that are relevant for extending them
to the ordinal case.

@@

%% Paragraph on test statistics
Specific test statistics commonly obtained under this framework
include the double maximum statistic
\begin{equation}
    \label{eq:dmax}
    \mathit{DM} = \max_{i = 1,\dots, n} \max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,
\end{equation}
which essentially tests whether any component of the cumulative score
process strays too far from zero and is easily visualized.  This test
discards information 
related to multiple parameters fluctuating simultaneously, resulting
in it having relatively low power for assessing measurement
invariance when multiple factor analysis parameters change
simultaneously \cite{MerZei13}.

@@

The more interesting result of Figure~\ref{fig:sim2res} lies in the comparison
of the four test statistics within each panel of the figure.  The
two ``unordered'' test statistics both have relatively high power to
detect the measurement invariance violation, illustrating the result 
of \citeA{Ben80} and others that the
likelihood ratio test statistic picks out minor 
parameter discrepancies at large $n$.  In contrast, the two ordinal
test statistics that we proposed have considerably lower power, with
the $\mathit{WDM}_o$ statistic being the lowest and the $\max\mathit{LM}_o$
statistic being higher at larger values of $d$.

@@

\subsection{Extension to Other Models}
We focused on testing for measurement invariance in factor
analysis models here, but the proposed test statistics are applicable to
other psychometric models that are estimated via ML (or similar estimation
techniques for independent observations that are governed by a central limit
theorem). The only
requirement for carrying out the tests is that the casewise scores
(Equation~\eqref{eq:score}) be available following model estimation.
As a result, applications to studying DIF in IRT are immediate (for a
presentation involving non-ordinal variables, see \citeNP{StrKop10}),
as are general psychometric applications to studying parameter
stability w.r.t.\ ordinal auxiliary variables.  These could 
include applications where ordinal variables are explicitly included
in the model, such as ordinal factor analysis.
We expect the
same general results to hold for these applications, whereby the
proposed test statistics are better than the LRT for detecting
monotonic instabilities.   The strucchange package \cite{Zei06} noted
previously can be used for these more-general applications.

\subsection{Summary}
As demonstrated via simulation, the proposed test statistics 
have relatively high power for detecting measurement 
invariance violations that are monotonic with the ordinal variable,
and they have relatively low power for detecting minor violations that are not
monotonic.  The former feature implies that the statistics are good at
detecting measurement invariance violations that are interpretable to
the researcher, while the latter feature implies that the statistics
are feasible in situations where the likelihood ratio test commonly
rejects $H_0$ in practice (e.g., \citeNP{Ben80}).  Furthermore, the
focal psychometric model does not have to be modified in any way,
which differs from approaches that may treat the ordinal variable as
continuous.  In all, the tests have
advantageous properties that should be useful in practice.
